%!TEX root = ../atomic.sigconf.tex
% LTeX: language=en
%
\section{Concluding Remarks}
\label{sec:conclusion}

We have given a sufficient condition for \kl{equivariant Gröbner bases} to be
computable, under natural \kl{computability assumptions}, and we have shown
that our sufficient condition is close to being optimal since the
undecidability of the \kl{equivariant ideal membership problem} can be derived
for a large class of group actions that do not satisfy our condition.
Let us now discuss some open questions and conjectures that arise from our work.

\para{Total orderings on the set of indeterminates} We assumed that the
indeterminates $\Indets$ were equipped with a total ordering $\varleq$ that is
preserved by the group action. This assumption seems necessary, as the notions
of \kl{leading monomials} would cease to be well-defined without it. However,
we do not have a clear understanding of whether this assumption is vacuous or
not. Indeed, as noticed by \cite[Lemma 13]{GHOLAS24}, and
\cref{thm:reducts-computable}, it often suffices to extend the structures of
the indeterminates to account for a total ordering. A conjecture of Pouzet
\cite[Problems 12]{POUZ24} states that such an ordering always exists, and this
was remarked by \cite[Remark 14]{GHOLAS24}. Note that in this case, one would
get a complete characterisation of the group actions for which the
\kl{equivariant Hilbert basis property} holds \cite[Property 4]{GHOLAS24}.

%\para{Labelled well-quasi-orderings and dichotomy conjectures} As noted in
%\Cref{sec:undecidability}, there are many conjectures relating the fact that
%$(\mon[Y]{\Indets}, \gdivleq)$ is a \kl{well-quasi-ordering} (for every
%\kl{well-quasi-ordered} set $Y$) and the presence of long
%paths of some kind (\cref{conj:wqo-infinite-path,rem:conj-wqo-infinite-path}).
%In particular, Pouzet's conjecture \cite{POUZ72} would imply that for actions
%arising from \kl{homogeneous structures} (as in the examples given in
%\cref{sec:act ex}), \cref{thm:compute-egb} and \cref{thm:undecidable-paths} are
%two sides of a dichotomy theorem: either the \kl{equivariant ideal membership
%problem} is undecidable and there are \kl{equivariant ideals} that are not
%\kl{orbit-finitely generated}, or every \kl{equivariant ideal} is
%\kl{orbit-finitely generated} and one can compute \kl{equivariant Gröbner
%bases}. Let us note that for some classes of graphs having bounded clique
%width, Pouzet's conjecture is known to hold \cite{DRT10,LOPEZ24}.
%This leads us to the following conjecture:
%
%\begin{conjecture}
%  For every action $\group\actson\Indets$ of a group $\group$ on a set
%  of indeterminates that is \kl{effectively oligomorphic}, exactly
%  one of the following holds:
%  \begin{enumerate}
%    \item The \kl{equivariant ideal membership problem} is decidable. 
%    \item There exists an \kl{equivariant ideal} that is not
%      \kl{orbit-finitely generated}.
%  \end{enumerate}
%\end{conjecture}
%
%Let us point out that a similar conjecture was already stated in the context of
%Petri nets with data. Indeed, the condition that $(\mon[Y]{\Indets},\gdivleq)$
%is a \kl{WQO} for every \kl{WQO} $Y$ also guarantees coverability of Petri nets
%with data $\X$ is decidable \cite[Theorem 1]{Lasota16}, and it was actually
%conjectured to be a necessary condition \cite[Conjecture 1]{Lasota16}. 

\para{Complexity} In the present paper, we have focused on the
decidability of the \kl{equivariant ideal membership problem} and the
computability of \kl{equivariant Gröbner bases}. However, we have not addressed
the complexity of such problems, and have only adapted the most basic
algorithms for computing \kl{Gröbner bases}. It would be interesting to know,
on the theoretical side, if one can obtain complexity lower bounds for such
problems, but also on the more practical side if advanced algorithms like
Faugère's algorithm \cite{FAUGERE02} can be adapted to the equivariant setting
and yield better performance in practice.


%\para*{Duals of orbit-finitely generated vector spaces} An interesting
%question posed by S\l{a}womir Lasota is that whether for group actions
%$\group\actson\Indets$ for which $(\mon[Y]{\Indets},\gdivleq)$ is a \kl{WQO}
%for every \kl{WQO} $Y$, the duals of orbit-finitely generated vector spaces are
%also orbit-finitely generated. This is a well-known fact that finitely
%generated vector spaces have finitely generated duals, and the extension to the
%orbit-finite setting would find applications in solving orbit-finite linear
%systems of equations and extend the results of \cite{GHL22}. It is already
%known to hold for actions originating from the \kl{homogeneous structures}
%considered in \Cref{ex:eq atoms,ex:dlo}, via the work of
%\cite{BFKM24,GHL22,Prz23}.

\para{Equivariant algebraic geometry} The development of \kl{equivariant
Gröbner bases} opens the door to the study of other classical results from
algebraic geometry to the equivariant setting. In particular, the status of the
fundamental result of algebraic geometry, \emph{Hilbert's Nullstellensatz},
that relates \kl{ideals} and \emph{varieties} (sets of common zeros of
polynomials), is still unclear in the equivariant setting. Other classical
notions, such as the one of Krull dimension, also deserve to be investigated
from the equivariant point of view.


