% LTeX: language=en
%%!TEX root = ../atomic.asmart.tex
%
\section{Relation to Existing Results and Examples}
\label{sec:examples}

In this section, we are interested in the consequences of our decidability
results. First, we will provide numerous examples of sets of indeterminates
that satisfy our \kl{computability assumptions} as well as our
\kl{well-quasi-ordering} conditions. Then, we will discuss how our results can
be applied to solve various decidability problems in theoretical computer
science.

\todo[inline]{for arka: integrate these examples}
\begin{example}
  \label{ex:q-is-super-wqo}
  The set $(\mon[Y]{\Indets_\Q}, \gdivleq)$ is a \kl{well-quasi-ordering} whenever $Y$ is
  one, and in particular $\Indets_\Q$ satisfies the computability assumptions and
  the termination assumptions of both \cref{thm:compute-egb}
  and
  \cref{thm:decide-equiv-ideal-mem}.
\end{example}
\begin{proof}
  Let $\seqof{\monelt_i}[i \in \N]$ be a sequence of monomials in
  $\mon[Y]{\Indets_\Q}$. Let us write each monomial $\monelt_i$ as
  a finite word $w_i$ over the alphabet $Y$, by writing all the exponents in the order 
  prescribed by the indeterminates.
  Because $Y$ is a \kl{well-quasi-ordering}, the set of all finite words over $Y$ is
  \kl{well-quasi-ordered} by the \emph{scattered subword} relation \cite{HIG52}.
  Now, if $w_i$ is a scattered subword of $w_j$, then
  $\monelt_i \gdivleq \monelt_j$, by choosing a suitable $\gelem \in \group$.
\end{proof}

\begin{example}
  \label{ex:z-is-not-wqo}
  The set $(\mon[Y]{\Indets_\Z}, \gdivleq)$ is not a \kl{well-quasi-ordering} whenever $Y$ is
  contains two distinct elements.
  In particular, $\Indets_\Z$ does not satisfy the termination assumptions of
  \cref{thm:compute-egb} and \cref{thm:decide-equiv-ideal-mem}.
\end{example}
\begin{proof}
  Assume that $Y$ has two distinct elements $a$ and $b$, and let us assume without loss of generality
  that $a \leq b$. The sequence of monomials 
  $\monelt_i \defined x_1^{b} x_2^{a} \cdots x_{i-1}^{a} x_i^{b}$
  forms an infinite antichain in $(\mon[Y]{\Indets_\Z}, \gdivleq)$.
  Indeed, if $\monelt_i \gdivleq \monelt_j$ for some $i < j$, then
  without loss of generality, $\gelem_i (x_1) = x_1$, and 
  $\gelem_i (x_i) = x_j$, because these are the only ones that can be 
  equipped with a large enough exponent.
  Therefore, $\gelem_i = \mathrm{id}$ since the group only contains translations.
  However, this implies that the exponent of $x_j$ in $\monelt_j$ is at most $a$,
  which contradicts the fact that it is $b$.
\end{proof}

\subsection{Crafting sets of indeterminates}
%
We give some interesting examples of group actions $\group \actson \Indets$ and discuss which of them satisfy the necessary condition of \Cref{thm:compute-egb}.
We also describe operations to build new group actions from old ones,
and discuss which of them preserve this condition.

\arka{Did we write somewhere that divisibility is same as labelled embedding}

\todo[inline]{for arka : add citations}

In all of our examples $\Indets$ is a set with some structure, described by some relations and functions on that set,
and $\group$ is the group $\aut{\Indets}$ of all automorphisms (i.e.\ bijections that preserve the structure) of $\Indets$.
To show that $(\mon[Y]{\Indets},\gdivleq)$ is a WQO we use the following strategy:
we define an equivariant one-to-one function $f_{\Indets} : \mon[Y]{\Indets} \to W_{\Indets}$ to some well-known well-quasi-ordered set $(W_{\Indets},\leq)$ such that $f_{\Indets}(\monelt[p]) \leq f_{\Indets}(\monelt[q])$ if and only if $\monelt[p] \gdivleq \monelt[q]$.
The reason why this strategy works is because in all of our examples, $\Indets$ will be a \intro{homogeneous} structure (\arka{cite wikipedia and macpherson survey}),
and $f_{\Indets}$ essentially maps an element $\monelt[p]$ of $\mon[Y]{\X}$,
thought as a finite induced substructure labelled by $Y$,
to its isomorphism class.
%
\begin{example}\label{ex:eq atoms}
Let $\A$ be an infinite set without any additional structure other than the equality relation.
Then $\aut{\A}$ is the set of all bijections of the set $\A$.
This action $\aut{\A} \actson \A$ does not preserve any linear order on $\A$.
However, $(\mon[Y]{\A}, \gdivleq)$ is a \kl{WQO} whenever $Y$ is a \kl{WQO}.
To see this define $f_{\A}$ to be the function which takes $\monelt[p]$ to the multiset of its coefficients.
For example, $f_{\A}(a^{y}a_2^{y'}c^{y} = \{y,y',y\}$ for every $a,b,c\in\A$ and $y,y'\in Y$.
We leave it to the reader to prove that for every $\monelt[p],\monelt[q]\in\mon[Y]{\A}$ we have $\monelt[p] \gdivleq[\aut{\A}] \monelt[q]$ if and only if $f(\monelt[p])$ is smaller than or equal to $f(\monelt[q])$ in the multiset ordering (\arka{cite MPRI lecture notes?}).
Note that the latter is a \kl{WQO}.
\end{example}
%
\begin{example}\label{ex:dlo}
Let $\D$ be a dense linear order without endpoints(\arka{cite wiki}).
Then $\aut{\D}$ is the set of all order preserving bijections of $\D$.
By definition, the action $\aut{\D} \actson \D$ preserves the linear order on $\D$.
Moreover, $(\mon[Y]{\A}, \gdivleq)$ is a \kl{WQO} whenever $Y$ is a \kl{WQO}.
Using the same strategy as \Cref{ex:eq atoms},
we define $f_{\D}$ to be the map which takes $\monelt[p]\in\mon[Y]{\D}$ to the corresponding word in $Y^*$ which we get by writing all the exponents in the order prescribed by the indeterminates.
For example, for every $a < b < c \in \D$ and $u,v\in Y$,
$f_{\D}(a^u b^v c^u) = uvu$.
We leave it to the reader to check that for every $\monelt[p],\monelt[q]\in\mon[Y]{\D}$ we have $\monelt[p] \gdivleq[\aut{\A}] \monelt[q]$ if and only if $f(\monelt[p])$ is smaller than or equal to $f(\monelt[q])$ in the scattered subword ordering, which is a \kl{WQO} due to Higman's lemma \arka{add citation}.
\end{example}
%
\begin{example}\label{ex:dlo}
Let $\calZ$ be the set of integers ordered by the usual ordering.
Then $\aut{\calZ}$ is the set of all order preserving bijections of $\D$.
Note that every order preserving bijection of the set $\calZ$ is a translation $n \mapsto n + c$ for some constant $c\in\calZ$.
By definition, the action $\aut{\calZ} \actson \calZ$ preserves the linear order on $\D$.
However, $(\mon[Y]{\calZ}, \gdivleq[\aut{\calZ}])$ is not a \kl{WQO} even when $Y$ is a singleton.
An example of an infinite antichain is the set $\setof{a b}{b\in\calZ\setminus\{a\}}$, for any fixed $a\in\calZ$.
\end{example}

\begin{example}\label{ex:rado}
Let $\G$ be the Rado graph (\arka{cite wiki}).
Then $\aut{\G}$ is the set of all automorphisms of the graph $\G$.
\end{example}
%
\begin{example}\label{ex:rado}
Let $\V$ be an infinite dimensional vector space over $\ftwo$.
Then $\aut{\V}$ is the set of all linear automorphisms.
\end{example}
%

\paragraph{Sets with atoms.}


\todo[inline]{Say that if one starts with atoms and equality, then we can only 
  have dimension 1, and that this is the case of the rationals.}

\paragraph{Relational structures.} Let $\mathbb{A}$ be an infinite relational
structure with finitely many relations. Then, one can consider the set of
polynomials $\poly{\K}{\mathbb{A}}$, where indeterminates are elements of the
universe of $\mathbb{A}$. The group of all automorphisms of $\mathbb{A}$ (i.e.,
bijections of the universe that preserve the relations) acts on
$\poly{\K}{\mathbb{A}}$ by permuting the indeterminates.

Natural examples are polynomials whose indeterminates are indexed by the
natural numbers (with inequality), or the rationals (with inequality). In this
setting, \kl{effective oligomorphcity} means that \todo{do it}. The fact that
$(\mon{\mathbb{A}}, \gdivleq)$ is a well-quasi-ordering corresponds to ordering
\emph{finite substructures} of $\mathbb{A}$ by the \emph{labelled induced
substructure} relation, and asking whether the class obtained is
well-quasi-ordered. This is a well-studied question in graph theory, where a
conjecture of Pouzet states that this holds with two labels if and only if it
holds for every ordinal. In particular, for such structures, it is therefore
conjectured that $(\mon{\Indets}, \gdivleq)$ is a well-quasi-ordering if and
only if $(\mon[\om \ordplus \ordfin{1}]{\Indets}, \gdivleq)$, $(\mon[\om
\ordplus \om]{\Indets}, \gdivleq)$, and $(\mon[\om^2]{\Indets}, \gdivleq)$ are
well-quasi-orderings too.
\arka{So we can use any well-ordered set of labels?} 
\todo[inline]{
  Cite \cite{POUZ72},
  \cite{DRT10} for the conjecture.
}


\paragraph{On reducts of structures.} \AP Let $\sigma$ be a finite relational
signature, and $\tau \subseteq \sigma$ be another finite relational signature.
Let $\mathbb{A}, \mathbb{B}$ be respectively a $\sigma$ and a $\tau$ structure. We say
that \intro{$\mathbb{B}$ is a reduct of $\mathbb{A}$} when $\mathbb{B}$ is
obtained from $\mathbb{A}$ by keeping the same universe, and relations. It was noted by \cite[Lemma 13]{GHOLAS24} that in
this case, the \kl{equivariant Hilbert basis property} transfers from
$\mathbb{A}$ to $\mathbb{B}$. Let us briefly argue that this transfer holds too
for our \cref{thm:decide-equiv-ideal-mem,thm:compute-egb}.

\begin{lemma}
  \label{lem:reducts-equiv-hilbert}
  Let $\mathbb{A}$ be a relational structure, let $\mathbb{B}$ be a 
  \kl(struct){reduct} of $\mathbb{A}$. Then, if $\mathbb{A}$ satisfies the
  hypotheses of \cref{thm:decide-equiv-ideal-mem},
  then one can decide the \kl{equivariant ideal membership problem} for
  $\poly{\K}{\mathbb{B}}$. Similarly, 
  if $\mathbb{A}$ satisfies the hypotheses of
  \cref{thm:compute-egb}, then one can compute an
  \kl{equivariant Gröbner basis} of an
  \kl{equivariant ideal} of $\poly{\K}{\mathbb{B}}$.
\end{lemma}
\begin{proof}
  \todo[inline]{Just write it, and it works.}
\end{proof}

\AP 
As a consequence, one can apply our results to structures that are not equipped 
with an ordering, because one can always consider the 

\todo[inline]{Talk about $\N$ I guess.}



\subsection{Topological Well-Structured Transition Systems}

The fact that (finite control) systems performing polynomial computations can
be verified folloms from the theory of \kl{Gröbner bases} on finitely many
indeterminates \cite{MULSEI02,BEDUSHWO17}. There were also numerous
applications to automata theory, suchz as deciding whether a weighted automaton
could be determinised (resp. desambiguated) \cite{BESM23,PUSM24}. We refer the
readers to a nice survey recapitulating the successes of the so-called
``Hilbert method'' automata theory \cite{BOJAN19}. In this section, we will
show how our results can be used in the framework of \kl{topological
well-structured transition systems}, that is an abstract setting to verify
infinite state systems that is more general than computing polynomial
invariants \cite{JGL07,JGL10}. As a consequence, we will show how our results
can be used to decide the zeroness of \kl{orbit finite polynomial automata}, a
new model of computation that generalizes \kl{polynomial automata}
\cite{BEDUSHWO17} and \kl{orbit finite weighted automata} \cite{BOKLMO21}.

\paragraph{Topological Well-Structured Transition Systems.} \AP The notion of
\kl{topological well-structured transition system} was introduced by
Goubault-Larrecq in \cite{JGL07}, noticing that the pre-existing notion of
\kl{Noetherian space} could serve as a topological generalisation of
\kl{well-quasi-orderings}, for which the celebrated decision procedures on
\kl{well-structured transition systems} can be applied. In particular,
Goubault-Larrecq used such systems to verify properties of \emph{polynomial
programs} computing over the rationals, that can communicate over lossy
channels using a finite alphabet \cite{JGL10}. 
\AP A \intro{topological space} is a set $X$ equipped with a collection $\tau$
of subsets of $X$ that is stable under finite intersections and arbitrary
unions.\footnote{In particular, $\tau$ contains the empty set and $X$ itself.}
In a \kl{topological space}, elements of $\tau$ are called \intro{open
subsets}, while their complements (in $X$) are called \intro{closed subsets}. A
\kl{topological space} is \intro(space){Noetherian} when, for every sequence
$\seqof{U_i}[i \in \N]$ of \kl{open subsets}, there exists $n \in \N$ such that
$\bigcup_{i \in \N} U_i = \bigcup_{i \leq n} U_i$. We refer the readers to the
book \cite{JGL13} for a comprehensive introduction to \kl{Noetherian spaces}
and their usage in theoretical computer science. Let us briefly argue that
\kl{Noetherian spaces} generalize \kl{well-quasi-orders} in
\cref{ex:well-quasi-orders-are-noeth}, and encode the
\kl{Hilbert basis property} in \cref{ex:polynomials-noetherian}.

\begin{example}[ see \cite{JGL13}]
  \label{ex:well-quasi-orders-are-noeth}
  Let $(X, \leq)$ be a quasi-ordered set.
  Then, the set $X$ equipped with the \kl{topology} having 
  as \kl{open subsets} the upwards-closed subsets of $X$ is \kl(space){Noetherian}
  if and only if $(X, \leq)$ is \kl{well-quasi-ordered}.
\end{example}

\begin{example}[ see \cite{JGL13}]
  \label{ex:polynomials-noetherian}
  Let $\K$ be a field, and let $n \in \N$.
  The space $\K^n$ equipped with the \kl{Zariski topology}
  \kl(space){Noetherian}; where the \intro{Zariski topology}
  is the topology whose \kl{closed subsets} are finite unions of sets
  of the form $\setof{ \vec{x} \in \K^n}{ \forall p \in \idl, p(\vec{x}) = 0}$,
  where $\idl$ is an \kl{ideal} of $\poly{\K}{x_1, \dots, x_n}$.
\end{example}

\AP The advantage of \kl{Noetherian spaces} over \kl{well-quasi-orderings} and
\kl{Noetherian rings} is that they generalize both and can be \emph{combined}:
\kl{Noetherian spaces} are closed under finite sums, finite products,
considering finite words, considering finite trees, and many more \todo{cite}.
As a consequence, they provide a versatile tool to express the set of states of
a system, ensuring that a strong termination property holds.

\AP A \intro{topological well-structured transition system} with alphabet
$\Sigma$ is a \kl{topological space} $(X, \tau)$, equipped with a transition
function $\delta \colon X \times \Sigma \to X$, such that the following
properties hold: for every $U \in \tau$, $\mathrm{pre}^\exists(U)$, the set of
states $x \in X$ such that there exists $a \in \Sigma$ with $\delta(x, a) \in
U$, is an \kl{open subset}. Equivalently, the set $\mathrm{pre}^\forall(E)$ of
states $x \in X$ such that for every $a \in \Sigma$, $\delta(x, a) \in E$ is a
\kl{closed subset} of $X$ whenever $E$ is itself a \kl{closed subset} of $X$.
The natural decition problem for \kl{topological well-structured transition
systems} is the following \intro{open reachability problem} is decidable: given
an initial state $x_0 \in X$ and an \kl{open subset} $U \in \tau$, is it true that
there exists a word $w \in \Sigma^*$ such that $\delta^*(x_0, w) \in U$? The
prototypical algorithm to solve this problem is the following \intro{backward
algorithm}: start with $U_0 \defined U$, and iteratively compute $U_{i+1}
\defined U_i \cup \mathrm{pre}^\exists(U_i)$ until $U_i = U_{i+1}$, then check
whether $x_0 \in U_\text{last}$.
There are easy-to-state sufficient conditions  for such an algorithm to be computable and terminate:
\begin{enumerate}
  \item One is equipped with an effective representation of open subsets,
    where one is able to test equality of open subsets, compute unions of open subsets, and test 
    membership of a point in an open subset.
  \item The pre-image function $\mathrm{pre}^\exists$ is computable, i.e., one can
    compute the set $\mathrm{pre}^\exists(U)$ for every open subset $U$.
  \item The space $(X, \tau)$ is \kl{Noetherian}. 
\end{enumerate}

\AP Our \cref{cor:equivariant-ideals-computations} shows that
under some assumptions on $\Indets$, the set of finitely supported functions
$\Indets \to \K$ is a \kl{Noetherian space} with respect to the
\intro{equivariant Zariski topology}, i.e., the topology whose \kl{closed subsets}
are finite unions of sets of the form $E_{\idl} \defined \setof{f \in
\K^{(\Indets)}}{\forall p \in \idl, p(f) = 0}$, where $\idl$ is an
\kl{equivariant ideal} of $\poly{\K}{\Indets}$. Furthermore, we have an
effective representation of the \kl{closed subsets} in this topology, using
\kl{equivariant Gröbner bases} of \kl{equivariant ideals}. In particular, the
theory of \kl{topological well-structured transition systems} can be applied to
systems whose state space contains ``named registers'' that contain numbers and
are updated by polynomial functions.


\paragraph{Consequences for orbit-finite polynomial automata.} Before
discussing the case of orbit finite polynomial automata, let us recall in
\cref{ex:polynomial-automata}
the
setting of \kl{polynomial automata} in the classical case, as studied by
\cite{BEDUSHWO17}, with techniques that dates back to \cite{MULSEI02}. We will
formally state in \cref{lem:zeroness-problem-polynomial-automata}
how the classical problem of deciding the \kl(pa){zeroness} of a
\kl{polynomial automaton} is a special case of the \kl{open reachability
problem} for \kl{topological well-structured transition systems}. Beware that
these are consequences of \cite[Section 6, Polynomial Games]{JGL10}.

\begin{definition}[Polynomial automata, as described in \cite{BEDUSHWO17}]
  \label{ex:polynomial-automata}
  A \intro{polynomial automaton} is a tuple $A \defined (Q, \Sigma, \delta, q_0, F)$,
  where $Q = \K^n$ for some finite $n \in \N$, $\Sigma$ is a finite alphabet,
  $\delta \colon Q \times \Sigma \to Q$ is a transition function such that 
  $\delta(\cdot,a)_i$ is a polynomial in the indeterminates $q_1, \dots, q_n$ for every
  $a \in \Sigma$ and every $i \in \set{1, \dots, n}$, $q_0 \in Q$ is the initial state,
  and $F \colon Q \to \K$ is a polynomial function describing the final result of the 
  automaton.
  The \intro{zeroness problem for polynomial automata} is the following decision problem:
  given a \kl{polynomial automaton} $A$, is it true that 
  for all words $w \in \Sigma^*$, the polynomial $F(\delta^*(q_0, w))$ is zero?
\end{definition}

\begin{lemma}
  \label{lem:zeroness-problem-polynomial-automata}
  The \kl{zeroness problem for polynomial automata} is a special case of the
  \kl{open reachability problem} for \kl{topological well-structured transition systems}.
\end{lemma}
\begin{proof}
  Let $A = (Q, \Sigma, \delta, q_0, F)$ be a \kl{polynomial automaton}.
  We consider the topological space $(Q, \tau)$, where $\tau$ is the
  \kl{Zariski topology} on $\K^n$.
  Let $\idl$ be an \kl{ideal} of $\poly{\K}{x_1,\dots,x_n}$ generated by the polynomials
  $p_1, \dots, p_m$,
  and let $E \defined \setof{q \in Q}{\forall p \in \idl, p(q) = 0}$,
  a \kl{closed subset} of $Q$.
  Then,
  \begin{align*}
    q \in \mathrm{pre}^\forall(E) & \iff 
    \forall a \in \Sigma, \forall p \in \idl, p(\delta(q, a)) = 0 \\
                                  & \iff 
    \forall a \in \Sigma, \forall p \in \idl, p(\delta(q, a)) = 0 \\
                                  & \iff 
                                  \forall p \in \idl[J], p(q) = 0
  \end{align*}
  where $\idl[J] \defined \IdlGen{ \setof{ p_i[ x_i \mapsto \delta(\cdot, a)_i] }{ i \in \set{1, \dots, m}, a \in \Sigma } }$.
  In particular, one can represent \kl{closed subsets} of $Q$ as finite 
  lists of \kl{ideals} using their \kl{Gröbner bases}, and we showed that 
  one can effectively compute the pre-image of \kl{closed subsets} of $Q$
  via $\mathrm{pre}^\forall$ by substituting polynomials.
  In this representation, it is very easy to compute the union 
  of two \kl{closed subsets}, which is simply concatenating the two lists 
  of \kl{ideals} reperesenting them.
  To compute the intersection of two \kl{closed subsets} $E_1$ and $E_2$,
  one can assume without loss of generality that both are represented by a 
  single ideal (i.e., that they are irreducible closed subsets), respectively 
  $\idl_1$ and $\idl_2$.
  Then, an easy computation shows that 
  $E_1 \cap E_2 = \setof{q \in Q}{\forall p \in \idl_1 + \idl_2, p(q) = 0}$,
  where $\idl_1 + \idl_2$ is the sum of the two ideals.
  Whether a point $q \in Q$ is in a \kl{closed subset} $E$ is decidable
  because one can evaluate the generating polynomials on $q$ and check that 
  it is indeed $0$.
  The equality check is more complicated, and can be done by first 
  normalizing the list of ideals so that their intersection is trivial,
  which requires computing the intersection of ideals
  and performing equality checks on the resulting \kl{ideals}.

  As a consequence, it suffices to test the \kl{open reachability problem} for
  the \kl{topological well-structured transition system} $(Q, \tau)$ with the
  initial state $q_0$ and the \kl{open subset} $U = Q \setminus E_\text{final}$,
  where $E_\text{final} \defined \setof{q \in Q}{F(q) = 0}$ is the \kl{closed subset}
  of states where the automaton outputs zero.
\end{proof}

\AP Let us fix a group $\group$ that acts on the set of indeterminates
$\Indets$, and on an alphabet $\Sigma$ in an \kl{effectively oligomorphic}
fashion. Let us now consider the case of \intro{orbit finite polynomial
automata}, that we define as follows: an \reintro{orbit finite polynomial
automaton} is a tuple $A \defined (Q, \Sigma, \delta, q_0, F)$, where $Q =
\K^{(\Indets)}$, $\Sigma$ is an \kl{orbit finite} alphabet, $\delta \colon
\Sigma \to (\Indets \to \poly{\K}{\Indets})$ is a \kl(func){finitely supported}
polynomial update function, and $F \in \poly{\K}{\Indets}$ is a polynomial
computing the result of the automaton. Given a letter $a \in \Sigma$ and a
state $q \in Q$, the update $\delta^*(q,a)$ is defined as the function from
$\Indets$ to $\K$ defined by $\delta^*(q,a) \colon x \mapsto \delta(a,x)[ q ]$,
which is well-defined because $\delta(a,x)$ is a \kl{finitely supported}
polynomial. The update function is naturally extended to words. Finally, the
output of an \kl{orbit finite polynomial automaton} on a word $w \in \Sigma^*$
is defined as $F(\delta^*(q_0, w))$.

\begin{example}
  \label{ex:orbit-finite-polynomial-automata}
  Let $\Indets = \Q$, and let $\group$ be the group of all
  order-preserving bijections of $\Q$.
  Let $\Sigma \defined \Q \times \Q$.
  Then, the following function are computable by 
  \kl{orbit finite polynomial automata}:
  the number $\mathrm{inc}(w)$ of letters $(a,b)$ such that $a < b$ in a word $w \in \Sigma^*$,
  the number $\mathrm{dec}(w)$ of letters $(a,b)$ such that $a > b$ in a word $w \in \Sigma^*$,
  and the number $(\mathrm{inc}(w) - \mathrm{dec}(w))^2$.
\end{example}

\AP As for \kl{polynomial automata}, the \intro(ofpa){zeroness problem} for
orbit finite polynomial automata is the following decision problem: decide if
for every input word $w$, the output $F(\delta^*(q_0, w))$ is zero. Solving the
\kl(ofpa){zeroness problem} for orbit finite polynomial automata allows us to
decide the equality of two such automata, by computing their substraction. 

It follows from the same reasoning as in
\cref{lem:zeroness-problem-polynomial-automata} that the
\kl(ofpa){zeroness problem} for orbit finite polynomial automata reduces to the
\kl{open reachability problem} for \kl{topological well-structured transition
systems} with the \kl{equivariant Zariski topology} on $\K^{(\Indets)}$.

\begin{corollary}
  \label{cor:orbit-finite-polynomial-automata-zeroness}
  Let $Y \defined (\N \ordplus \ordfin{1})^2$ be the set of pairs 
  of potentially infinite natural numbers ordered pointwise.
  Let $\Indets$ be a set of indeterminates that satisfies the
  \kl{computability assumptions} and such that $(\mon[Y]{\Indets}, \gdivleq)$ is a
  \kl{well-quasi-ordering}.
  Then, the \kl(ofpa){zeroness problem} is decidable for all \kl{orbit finite polynomial automata}
  with register names in $\Indets$, for every \kl{orbit finite} alphabet $\Sigma$,
  that is \kl{effectively oligomorphic} with respect to the action of $\group$.
\end{corollary}
\begin{proof}
  Let $A = (Q, \Sigma, \delta, q_0, F)$ be an \kl{orbit finite polynomial automaton}.
  We consider the topological space $(Q, \tau)$, where $\tau$ is the
  \kl{equivariant Zariski topology} on $\K^{(\Indets)}$.
  Let $E_\text{final} \defined \setof{q \in Q}{F(q) = 0}$ be the \kl{closed subset}
  of states where the automaton outputs zero.

  We proceed as in \cref{lem:zeroness-problem-polynomial-automata},
  representing \kl{closed subsets} of $Q$ as finite lists of \kl{equivariant
  ideals}, which is an effective representation of \kl{closed subsets} thanks
  to \cref{cor:equivariant-ideals-computations}. The only
  non-trivial check to apply the same reasoning as in
  \cref{lem:zeroness-problem-polynomial-automata} is that
  $\mathrm{pre}^\forall(E)$ is computable for every \kl{closed subset} $E$ of
  $Q$. To that end, let us fix an \kl{equivariant ideal} $\idl$ and an
  \kl{orbit finite} set $H$ of polynomials such that $\idl = \IdlGen{H}$. Then,
  we can compute the \kl{equivariant ideal} $\idl[J]$ generated by the
  polynomials $\mathrm{pullback}(p,a) \defined p [ x_i \mapsto \delta(a)(x_i)]$ for every pair $(p, a) \in H \times \Sigma$. Indeed, $H \times \Sigma$
  remains \kl{orbit finite} because the action of $\group$ on $\Indets$ is
  \kl{effectively oligomorphic}, and the function $\mathrm{pullback}$ is computable
  and \kl(func){equivariant}: indeed, given  $\gelem \in \group$, we can show that
  \begin{align*}
    \gelem \cdot \mathrm{pullback}(p, a) & = 
    \gelem \cdot (p [ x_i \mapsto \delta(a)(x_i)]) \\
    & = p [ x_i \mapsto (\gelem \cdot \delta(a, x_i))] \\
    & = p [ x_i \mapsto \delta(\gelem \cdot a, \gelem \cdot x_i))] \\
    & = (\gelem \cdot p) [ x_i \mapsto \delta(\gelem \cdot a, x_i)] \\
    & = \mathrm{pullback}(\gelem \cdot p, \gelem \cdot a).
  \end{align*}
  As a consequence, we can 
  compute an \kl{orbit finite set} of polynomials $H'$ that generate $\idl[J]$.
  Furthermore, it is clear that 
  $\mathrm{pre}^\forall(E)$ is exactly the set of states $q \in Q$ such that
  $\forall p \in \idl[J], p(q) = 0$.
  We have concluded.
\end{proof}

While all of these reasosing could be done outside the realm of (effective)
\kl{topological well-structured transition systems}, we can use the modularity
of the theory to obtain more complex verification properties. Following the
lines of \cite[Theorem 6]{JGL10}, one can consider the case of communicating
orbit finite polynomial automata, where we have a collection processes that
communicate letters over a finite alphabet using lossy channels, and can
perform polynomial updates on their local state. Deciding whether such a system
can reach a state where one process fails to satisfy a given polynomial
invariant is a special case of the \kl{open reachability problem}, and is
decidable.


\paragraph{Reachability problem of symmetric data Petri nets.}

\arka{To add : symmetric VAS equations}
\todo[inline]{for arka: define symmetric data Petri nets/ symmetric VAS equations}

\paragraph{Orbit-finite systems of equations}

\todo[inline]{for arka: do it}

