%!TEX root = ../atomic_lipics.tex
% LTeX: language=en
\section{Preliminaries}
\label{sec:preliminaries}

%\subsection{Well-quasi-orders}
\subsection{Partial orders, ordinals, well-founded sets, and well-quasi-ordered sets.}
\AP We assume basic familiarity with partial orders, well-founded sets,
and ordinals. We use the notation $\intro*\om$ for the first infinite
ordinal (that is, $(\N, \leq)$), and write $X \intro*\ordplus Y$ for the
lexicographic sum of two partial orders $X$ and $Y$. Similarly, the notation $X
\intro*\ordtimes Y$ denotes the product of two partial orders equipped with the
lexigographic ordering, i.e. $(x_1, y_1) \leq (x_2, y_2)$ if either $x_1 <
x_2$, or $x_1 = x_2$ and $y_1 \leq y_2$. We also use the usual notations
for finite ordinals, writing $\intro*\ordfin{n}$ for the finite ordinal of size
$n$. For instance, $\om \ordplus \ordfin{1}$ is the total order $\N \uplus
\set{+\infty}$, where $+\infty$ is the new largest element.

\AP In order to guarantee the termination of the algorithms presented in this
paper, a key ingredient is the notion of \intro{well-quasi-ordering}
(WQO), which consists of sets $(X, \leq)$ such that every infinite sequence
$\seqof{x_i}[i \in \N]$ of elements of $X$ contains a pair $i < j$ such that
$x_i \leq x_j$. Examples of \kl{well-quasi-orderings} include finite sets with
any ordering, or $\N \times \N$ with the product ordering. We refer the reader
to \cite{SCSC17} for a comprehensive introduction to \kl{well-quasi-orderings}
and their applications in computer science.

\subsection{Polynomials, monomials, divisibility.} \AP 
We assume basic familiarity with the theory of
commutative algebra and polynomials. We use the notation $\poly{\K}{\Indets}$
for the ring of polynomials with coefficients from a field $\K$ and
indeterminates from a set $\Indets$, and $\mon{\Indets}$ for the set of
monomials in $\poly{\K}{\Indets}$. Letters $p,q,r$ denote polynomials,
$\monelt,\monelt[n]$ denote monomials, and $a,b,\alpha,\beta$ denote coefficients in $\K$.

A classical example of a \kl{WQO} is the set of monomials $\mon{\Indets}$,
endowed with the \kl{divisibility} relation $\intro*\divleq$ when $\Indets$
is finite. We recall that a monomial $\monelt[m]$ \intro{divides} a monomial
$\monelt[n]$ if there exists a monomial $\monelt[l]$ such that $\monelt[m]
\times \monelt[l] = \monelt[n]$. In this case, we write $\monelt[m]
\reintro*\divleq \monelt[n]$. Note that monomials can be seen as functions from
$\Indets$ to $\N$ with finite support, and that the \kl{divisibility}
relation can be extended to monomials that are functions from $\Indets$ to
$(Y,\leq)$, where $Y$ is any partially ordered set. In this case, we write
$\monelt[m] \divleq \monelt[n]$ if for every $x \in \Indets$, we have
$\monelt[m](x) \leq \monelt[n](x)$. We write $\intro*\mon[\om \ordplus
1]{\Indets}$ (resp. $\mon[\om^2]{\Indets}$) for the set of monomials that are
functions from $\Indets$ to $\om \ordplus \ordfin{1}$ (resp. $\om^2$).

\arka{example here? why not $\mon[Y]{\Indets}$?}

\AP Unless otherwise specified, we assume that the set of indeterminates
$\Indets$ comes equipped with a total ordering $\varleq$. Using this order, we
define the \intro{reverse lexicographic} (revlex) ordering on monomials as
follows: $\monelt[n] \intro*\revlexlt \monelt[m]$ if there exists an
indeterminate $x \in \Indets$ such that $\monelt[n](x) < \monelt[m](x)$, and such
that for every $y \in \Indets$, if $x \varlt y$ then $\monelt[n](y) =
\monelt[m](y)$. Note that if $\monelt[n] \divleq \monelt[m]$, then in
particular $\monelt[n] \revlexleq \monelt[m]$. 

\arka{example here?}
\arka{remark that reverse is important and point to where we explain why}

\AP We can now use the \kl{reverse lexicographic} ordering to identify particular elements in
a given polynomial. Namely, for a polynomial $p \in \poly{\K}{\X}$, we define
the \intro{leading monomial} $\intro*\lm(p)$ of $p$ as the largest monomial
appearing in $p$ with respect to the \kl{revlex ordering}, and the
\intro{leading coefficient} $\intro*\lc(p)$ of $p$ as the coefficient of
$\lm(p)$ in $p$. We then define the \intro{leading term} $\intro*\lt(p)$ of
$p$ as the product of its \kl{leading monomial} and its \kl{leading
coefficient}, and the \intro{characteristic monomial} $\intro*\cm(p)$ of $p$ as
the product of its \kl{leading monomial} and all the indeterminates appearing
in $p$. We also define the \intro(monomial){domain} of $\monelt[m]$ as the set
$\intro*\dom(\monelt[m])$ of indeterminates $x \in \X$ such that $\monelt[m](x) \neq
0$. Because the coefficients and monomial in question are highly dependent on
the ordering $\varleq$, we allow ourselves to write $\lm[\Indets](p)$ to
highlight the precise ordered set of variables that was used to compute the
\kl{leading monomial} of $p$. We extend $\dom$ from 
monomials to polynomials by defining $\dom(p)$ as 
the union of the \kl(monomial){domains} of all monomials appearing in $p$.
\arka{examples}


\arka{push this remark later and make it more concrete}
\begin{remark}\label{rem:other-monomial-orderings}
In the case of a finite set of indeterminates, one can choose any total ordering
on $\mon{\Indets}$, as long as it contains the \kl{divisibility}
quasi-ordering, and is compatible with the product of monomials.\footnote{This
is often called a \emph{monomial ordering}, see \cite{CLO15}.} In our case,
having an infinite number of indeterminates, we rely on a connection between
$\lm(p)$ and $\dom(p)$: $\dom(p) \subseteq \dwset{\dom(\lm(p))}$, where
$\dwset{S}$ is the downwards closure of a set $S \subseteq \Indets$, i.e. the
set of all indeterminates $x \in \Indets$ such that $y \leq x$ for some $y \in
S$. This means that the \kl{leading monomial} encodes a \emph{global property}
of the polynomial, and it will be crucial in our termination arguments. This is
already at the core of the classical \emph{elimination theorems} \cite[Chapter 3, Theorem
2]{CLO15}.
\end{remark}

%\subsection{Ideals and Gr\"{o}bner Bases}
\subsection{Ideals, and Gröbner Bases.}
\AP An \intro{ideal} $\idl$ of
$\poly{\K}{\X}$ is a non-empty subset of $\poly{\K}{\X}$ that is closed under
addition and multiplication by elements of $\poly{\K}{\X}$. Given a set $H
\subseteq \poly{\K}{\X}$, we denote by $\intro*\IdlGen{H}$ the ideal generated
by $H$, i.e. the smallest ideal that contains $H$. The \intro{ideal membership
problem} is the following decision problem: given a polynomial $p \in
\poly{\K}{\X}$ and a set of polynomials $H \subseteq \poly{\K}{\X}$, decide
whether $p$ belongs to the ideal $\IdlGen{H}$ generated by $H$. We know that
this problem is decidable when $\X$ is finite, and that it is even
$\EXPTIME$-complete \cite{MAME82}. The classical approach to the \kl{ideal
membership problem} is to use the \kl{Gröbner basis} theory, developed
in the 1970s by Buchberger~\cite{BUCH76}. 
A set $\Basis$ of polynomials is called a \intro{Gröbner basis} of
an ideal $\idl$ if, $\IdlGen{\Basis} = \idl$ and for every polynomial $p \in
\idl$, there exists a polynomial $q \in \Basis$ such that $\lm[\Indets](q)
\divleq \lm[\Indets](p)$.

Given a \kl{Gröbner basis} $\Basis$ of an ideal $\idl$, and a polynomial $p$,
it suffices to iteratively reduce the \kl{leading monomial} of $p$ by
subtracting multiples of elements in $\Basis$, until one cannot apply any
reductions. If the result is $0$, then $p$ belongs to $\idl$, and otherwise it
does not. 

\begin{example}
  Let $\Indets \defined \set{x,y,z}$ with $z < y < x$. The
  set $\Basis \defined \set{x^2y - z, x^2 - y}$ is not 
  a \kl{Gröbner basis} of the ideal $\idl$ it generates,
  because the polynomial $p \defined y^2 - z$ belongs to $\idl$
  but its \kl{leading monomial} $y^2$ is not divisible by 
  $\lm(x^2y - z) = x^2y$ nor by $\lm(x^2 - y) = x^2$.
\end{example}

\subsection{Group actions and equivariance} 
%
In this article we are interested in polynomial rings whose variables are identified with elements of a \kl{countable relational structure}.

A \intro{countable relational structure} $\Indets = (X,R_1,R_2,\dots,f_1,f_2,\dots)$ is given by a countable set $X$ with relations $R_i \subseteq X^{m_i}$ and functions $f_j : X^{n_j}\to X$ defined on it.
By an \intro{element} of the relational structure $\Indets$,
we mean an element of the underlying set $X$.
Similarly, we use $\Indets^n$ to denote the set $X^n$.

To illustrate our definitions and results we will mainly use two following relational structures:
%
\begin{enumerate}
\item The relational structure $\EqualityAtoms = (\Q,=)$ of rational numbers with equality.
\item The relational structure $\OrderAtoms = (\Q,<)$ of rational numbers with order.
\end{enumerate}
%
More examples of relational structures can be found in \arka{cite}
%
An \intro{automorphism} $\pi$ of $\Indets$ is a bijection of the underlying set that preserves and reflects all relations and commutes with all functions:
\[
R_i(x_1,\dots,x_{m_i}) \iff
R_i(\pi(x_1),\dots,\pi(x_{m_i}))
\text{ for all } (x_1,\dots,x_{m_i})\in \Indets^{m_i}, \text{ and}
\]
%
\[
\pi(f_j(x_1,\dots,x_{n_j})) \iff
f_j(\pi(x_1),\dots,\pi(x_{m_i}))
\text{ for all } (x_1,\dots,x_{n_j})\in \Indets^{n_j}.
\]
%
We use \intro{$\aut{\Indets}$} to denote the set of automorphisms of the relational structure $\Indets$.
%
\begin{example}
An automorphism of $\EqualityAtoms$ is simply a bijection from the set $\N$ to itself.
An automorphism of $\OrderAtoms$ is an order preserving bijection from $\Q$ to itself.
\end{example}
%
We use \intro{$\poly{\K}{\X}$} to denote the polynomial ring over $\K$ which uses elements of $\X$ as variables.
A subset $S \subseteq \poly{\K}{\Indets}$ is equivariant if it is invariant under automorphisms of $\Indets$.
%
%\begin{example}
%Both ideals $\idl[I]_0^{\EqualityAtoms}$ and $\idl[I]_1^{\EqualityAtoms}$ are equivariant.
%\end{example}
%
\begin{example}
    \label{ex:idl-equiv}
    Then the set $S_0 \subset \poly{\K}{\EqualityAtoms}$ of all polynomials 
    whose set of coefficients sums to $0$ is an equivariant ideal.
    Conversely, the set of all polynomials that are multiple
    of $x \in \EqualityAtoms$ is an \kl{ideal} that is not \kl{equivariant}.
\end{example}
%
\begin{proof}
    Let $p,q\in S_0$, and $r \in \poly{\K}{\EqualityAtoms}$.
    Then, $p \times r + q$ is in $S_0$. Remark that 
    $p,r$ and $q$ belong to a subset $\poly{\K}{\EqualityAtoms}$ of the 
    polynomials that uses only finitely many indeterminates.
    In this subset, the sum of all coefficients is obtained
    by applying the polynomials to the value $1$ for every indeterminate
    $y \in \EqualityAtoms$. We conclude that
    $(p \times r + q)(1,\dots, 1) 
    = p(1,\dots,1) \times r(1,\dots,1) + q(1,\dots,1)
    = 0 \times r(1, \dots, 1) + 0 = 0$, hence that
    $p \times r + q$ belongs to $S_0$. 
    Because $0$ is in $S_0$, we conclude that $S_0$ is an \kl{ideal}.
    Furthermore, if $\gelem \in \group$ and $p \in S_0$, then
    the sum of the coefficients $\gelem \cdot p$ is exactly
    the sum of the coefficients of $p$, hence is $0$ too.
    This shows that $S_0$ is \kl(ideal){equivariant}.

    It is clear that all multiples of a given polynomial $x \in \EqualityAtoms$
    is an ideal of $\poly{\K}{\EqualityAtoms}$.
    This is not an \kl{equivariant ideal}:
    take any bijection $\gelem \in \aut{\EqualityAtoms}$ that does not map
    $x$ to $x$ (it exists because $\EqualityAtoms$ is infinite and
    $\aut{\EqualityAtoms}$ is all permutations), then $\gelem(x)$ is not a 
    multiple of $x$, and therefore does not belong to the ideal.
\end{proof}
%
\subsection{Orbit finiteness}
%
%For computational purposes we will represent objects (monomials, polynomials, etc) constructed from $\Indets$ as pairs $(\vec{x},s)$ where $\vec{x} \in\Indets^*$ and $s$ is a object (or more formally a set) constructed without using $\Indets$.
%For instance, the polynomials $(a_1 + a^2_2), (a_1 + a_3^2) \in \poly{\K}{\EqualityAtoms}$ are represented as pairs $(a_1a_2,p), (a_1a_3,p)$ where $p$ is the polynomial $x_1 + x_2^2$.
\AP An \kl{set $S$} is said to be \intro{orbit finite} if it is the union of finitely many \intro{orbits} under the action of $\aut{\Indets}$.
We denote $\intro*\orbit{E}$ for the set of all elements $\gelem(x)$ for $\gelem \in \aut{\Indets}$ and $x \in E$. Equivalently, an \reintro{orbit finite set}
is a set of the form $\orbit{E}$ for some finite set $E\subseteq \poly{\K}{\Indets}$.
For instance the orbit of the polynomial $(x - 1)\in\poly{\K}{\EqualityAtoms}$ is the set $\setof{y - 1}{y\in\EqualityAtoms}$.

%Not every \kl{equivariant subset} is \kl{orbit finite}, as shown in \cref{ex:orbit-finite}.
%However, \kl{orbit finite sets} are
%robust in the sense that \kl{equivariant subsets} of \kl{orbit finite sets} are
%also \kl{orbit finite}, and similarly, an \kl{equivariant subset} of $E^n$ is
%\kl{orbit finite} whenever $E$ is \kl{orbit finite} and $n \in \N$ is finite.
%\arka{only for oligomorphic}
For algorithmic purposes, \kl{orbit finite sets} are those that can be taken
as input as a finite set of representatives (one for each orbit). The notions
of \kl{equivariance} and \kl{orbit finite sets} from a computational
perspective are discussed in \cite{BOJAN16inf}, and we refer the reader to this
book for a more comprehensive introduction to the topic.

\AP We will mostly be interested in \intro{orbit-finitely generated}
\kl{equivariant ideals}, i.e.\ \kl{equivariant ideals} that are generated by an
\kl{orbit finite set} of polynomials, for which the \intro{equivariant ideal
membership problem} is as follows: given a polynomial $p \in
\poly{\K}{\Indets}$ and an \kl{orbit finite set} $H \subseteq
\poly{\K}{\Indets}$, decide whether $p$ belongs to the \kl{equivariant ideal}
$\intro*\EqIdlGen{H}$ generated by $H$.

%\begin{example}
%  \label{ex:orbit-finite}
%  Let $\Indets = \N$, and $\group$ be all permutations 
%  that fixes prime numbers. The
%  set of all polynomials  whose coefficients sum to $0$ is an 
%  \kl{equivariant ideal}, but it is not \kl{orbit finite},
%  since all the polynomials $x_p - x_q$ for $p \neq q$ primes
%  are in distinct orbits under the action of $\group$.
%\end{example}

\AP A function $f \colon X \to Y$ between two sets $X$ and $Y$ equipped with
actions $\group \actson X$ and $\group \actson Y$ is said to be
\intro(func){equivariant} if for all $\gelem \in \group$ and $x \in X$, we have
$f(\gelem \cdot x) = \gelem \cdot f(x)$. For instance, the
\kl(monomial){domain} of a monomial is an \kl{equivariant function}: if $\gelem
\in \group$, then $\gelem \cdot \dom(\monelt[m]) = \dom(\gelem \cdot
\monelt[m])$. Let us point out that the image of an \kl{orbit finite set} under
an \kl{equivariant function} is \kl{orbit finite}, and that the algorithms that
we will develop in this paper are all \kl(func){equivariant}.
%
\subsection{Computability assumptions.}
%
We say that $\Indets$ is \intro{effectively oligomorphic} if:
%
\begin{enumerate}
\item It is \intro{oligomorphic}, i.e.\ for every $n \in \N$,
$\Indets^n$ is \kl{orbit finite},
\item There exists an algorithm that decides whether two elements $\vec{x},
\vec{y} \in \Indets^n$ are in the same orbit under the action of 
$\group$ on $\Indets^n$.
\item There exists an algorithm which, on input $n\in\N$, outputs a set
$A \subseteqfin \Indets^n$ such that $|A\cap U| = 1$ for every orbit
$U \subseteq \Indets^n$.
\end{enumerate}
%In particular, $\Indets$ itself is \kl{orbit finite} under the action of $\group$.
The structure $\Indets$ is called \intro{ordered} if one of the relations of $\Indets$ is a total order.
Let us point out that in this case,
$\revlexleq$ is equivariant, i.e. for all $\gelem \in \aut{\Indets}$ and
monomials $\monelt[m], \monelt[n] \in \mon{\Indets}$,
we have $\monelt[m] \revlexleq \monelt[n]$ if and only if $\gelem(\monelt[m]) \revlexleq \gelem(\monelt[n])$.
Our \intro{computability assumptions} on the tuple $\Indets$ is that it is \kl{ordered} and \kl{effectively oligomorphic}.
%
\begin{example}
  \label{ex:computability-assumptions}
  Both $\EqualityAtoms$ and $\OrderAtoms$ are \kl{effectively oligomorphic}. 
\end{example}
%
%Note that under our \kl{computability assumptions}, the set of polynomials
%$\poly{\K}{\Indets}$ is also \kl{effectively oligomorphic} under the action of
%$\group$ on $\Indets$ when restricted to polynomials with bounded degree.
%This is because a polynomial $p \in \poly{\K}{\Indets}$
%can be seen as an element of $(\K \times \Indets^{\leq d})^n$ where $n$ is the
%number of monomials in $p$, and $d$ is the maximal degree of a monomial
%appearing in $p$. Note that the set of all polynomials $\poly{\K}{\Indets}$ is
%not \kl{orbit finite}, precisely because the orbit of a polynomial $p$ under 
%the action of $\group$ cannot change the degree of $p$, and because there are 
%polynomials of arbitrarily large degree.
%\subsection{Equivariant Gröbner bases}
\subsection{Equivariant Gröbner bases.}
\AP
We know from \cite{GHOLAS24} that a
necessary condition for the \kl{equivariant Hilbert basis property} to hold is
that the set  $\mon{\X}$ of monomials is a \kl{well-quasi-ordering} when
endowed with the \intro{divisibility up-to equivariance} relation
($\intro*\eqdivleq$), which is defined as follows: for $\monelt_1,
\monelt_2 \in \mon{\Indets}$,
we write $\monelt_1 \gdivleq \monelt_2$ if there exists $\gelem \in \aut{\Indets}$ such that $\gelem(\monelt_1)$ \kl{divides} $\monelt_2$.

Let $\mon[Y]{\Indets}$ denote the set of all partial functions from $\Indets$ to a partially ordered set $(Y,<)$ with finite domain.
The reason behind the notation is that $\mon{\Indets}$ can be identified with $\mon[\N]{\Indets}$.
The \reintro{divisibility up-to equivariance} relation is extended to monomials as: for $\monelt_1,\monelt_2 \in \mon[Y]{\Indets}$,
we write $\monelt_1 \gdivleq \monelt_2$ if there exists $\gelem \in \aut{\Indets}$ such that the domain of $\gelem(\monelt[m]_1)$ is a subset of the domain of $\monelt[m]_2$ and $\gelem(\monelt[m]_1)(x) \leq \monelt[m]_2(x)$ for all $x$ in the domain of $\monelt[m]_1$.

Let $\Indets$ be a ordered relational structure and denote the ordering by $<$.
We say that a set $\Basis \subseteq \poly{\K}{\X}$ is an
\intro{equivariant Gröbner basis} of an equivariant ideal $\idl$ if $\Basis$ is \kl{equivariant},
$\IdlGen{\Basis} = \idl$, and for every polynomial $p \in \idl$, there exists
$q \in \Basis$ such that $\lm[<](q) \eqdivleq \lm[<](p)$ and
$\dom(q) \subseteq \dom(p)$, following the ideas of \cite{GHOLAS24}.

\AP We say that $\Indets$ is \intro{well-structured} if the set $(\mon[Y]{\Indets}, \eqdivleq)$ is \kl{well-quasi-ordered},
for every \kl{well-quasi-order} $(Y,\leq)$.
We say that it is \intro{$\omega$-well-structured} if $(\mon{\Indets}, \eqdivleq)$ is \kl{well-quasi-ordered}.

Note that even in the case of a finite set of variables, a \kl{Gröbner basis}
is not necessarily an \kl{equivariant Gröbner basis}, because of the
\kl(polynomial){domain} condition. However, every \kl{equivariant Gröbner
basis} is a \kl{Gröbner basis}.
%
\begin{example}
The set $\setof{x - 1}{x\in\OrderAtoms}$ is a \kl{equivariant Gröbner
basis} for the ideal $\idl[I]_1^{\OrderAtoms}$.
The set $\setof{x + y - 2}{x < y\in\OrderAtoms}$ is not a \kl{equivariant Gröbner
basis} for the ideal $\idl[I]_1^{\OrderAtoms}$.
Observe that $\setof{x + y - 2}{x < y\in\OrderAtoms}$ generates $\idl[I]_1^{\OrderAtoms}$ since for every $x < y < z$ we have $(x + y - 2) + (x + z - 2) - (y + z -2) = 2x - 2$.
\end{example}
%
%
%\begin{example}
%  \label{ex:equivariant-gb}
%  Let $\Indets \defined \set{ x_1, x_2 }$,
%  with $x_1 \varleq x_2$,
%  and $\group$ be the trivial group.
%  Let us furthermore consider the ideal $\idl$ \kl(idl){generated by}
%  $\set{ x_1, x_2 }$.
%  Then, the set $\Basis \defined \set{ x_2 - x_1, x_1 }$ is a
%  \kl{Gröbner basis} of $\idl$, but not an \kl{equivariant Gröbner basis}.
%  Indeed, $x_2 \in \idl$, but there is no polynomial $q \in \Basis$
%  such that $\lm(q) \divleq x_2$ and $\dom(q) \subseteq \dom(x_2)$.
%\end{example}
%
%
%\arka{Does not satisfy the domain condition}
%In the finite case, one can always compute an \kl{equivariant Gröbner basis} by
%computing \kl{Gröbner bases} for every possible ordering of the indeterminates,
%and taking their union.\footnote{This algorithm is correct because we are
%  considering the \kl{reverse lexicographic} ordering.}
 




