% LTeX: language=en
\section{Weak Equivariant Gröbner Bases}
\label{sec:weakgb}

\AP In this section we prove that a natural adaptation of \kl{Buchberger's
algorithm} to the equivariant setting computes a \kl{weak equivariant Gröbner
basis} of an \kl{equivariant ideal}. This can be seen as an analysis of the
classical algorithm in the equivariant setting. We will assume for the rest of
the section that $\X$ is a set of indeterminates equipped with a group $\group$
acting \kl{effectively oligomorphically} on $\X$, and that $\X$ is equipped
with a total ordering $\varleq$ that is \kl{compatible} with the action of
$\group$.

\begin{definition}
A \intro{weak equivariant Gröbner basis} of an \kl{equivariant ideal}
$\idl$ for an ordering $\preceq$ on polynomials is a finite set $\Basis$ of
polynomials such that for every polynomial 
\begin{enumerate}
  \item $\Basis$ is a generating set of an equivariant ideal $\idl$,
  \item For all $p \in \EqIdlGen{\Basis}$, there
    exists a polynomial $q \in \orbit[\group]{\Basis}$, an element $\gelem \in \group$,
    such that $p \toeucl{\pmonleq}{q} p'$, and
    $p'$ is a polynomial that only contains indeterminates that are smaller or
    equal than the indeterminates of $p$ with respect to the ordering $\varleq$.
\end{enumerate}
\end{definition}

In particular, we do not require that the rewriting relation 
is \kl{terminating} or \kl{confluent}, and we weaken 
the assumptions on the relationship between the rewriting relation
and the equivariant ideal generated by $\Basis$.

\AP Let us now construct a quasi-ordering $\preceq$ on polynomails based on the
ordering $\varleq$ on indeterminates. Traditionally, one first considers a
linearisation of the \kl{divisibility ordering} $\divleq$ on monomials, and
then extends it to a quasi-ordering on polynomials. Because we are working in
an equivariant setting, we will use a linearisation of $\gdivleq$ instead, but
unlike the classical case, the extension to polynomials will add another
ingredient.

\AP We will use the \kl{revlex} ordering on monomials, which is defined as
follows: $\monelt[n] \revlexlt \monelt[m]$ if there exists an indeterminate $x
\in \X$ such that $\monelt[n](x) < \monelt[m](x)$, and such that for every $y
\in \X$, if $x \varlt y$ then $\monelt[n](y) = \monelt[m](y)$. Remark that if
$\monelt[n] \revlexleq \monelt[m]$, then in particular $\monelt[n] \divleq
\monelt[m]$. This defines a total ordering on monomials that is \kl{compatible}
with the action of $\group$: if $\monelt[n] \revlexlt \monelt[m]$, then for
every $g \in \group$, we have $g \cdot \monelt[n] \revlexlt g \cdot
\monelt[m]$, precisely because the ordering $\varleq$ is \kl{compatible} with
the action of $\group$.

\AP We can now use the \kl{revlex} ordering to identify particular elements in
a given polynomial. Namely, for a polynomial $p \in \poly{\K}{\X}$, we define
the \intro{leading monomial} $\lm(p)$ of $p$ as the largest monomial appearing
in $p$ with respect to the \kl{revlex} ordering, and the \intro{leading
coefficient} $\lc(p)$ of $p$ as the coefficient of $\lm(p)$ in $p$. We can then
define the \intro{leading term} $\lt(p)$ of $p$ as the product of its
\kl{leading monomial} and its \intro{leading coefficient}, and the
\intro{characteristic monomial} $\cm(p)$ of $p$ as the product of its
\kl{leading monomial} and all the indeterminates appearing in $p$. 


\AP Using the notion of \kl{leading monomial}, we can define a quasi-ordering
on polynomials as follows: 
\begin{equation}
  \label{eq:chm}
  p \pmonlt q \iff
  \mathop{\land}
  \begin{cases}
    \lm(p) \revlexlt \lm(q) \\
    \dom(p) \subseteq \dom(q) \\
  \end{cases}
\end{equation}

Let us immediately notice a useful property of characteristic monomials
that will be useful throughout the paper. 

\begin{lemma}
  \label{lem:chm-compat}
  Let $p,q \in \poly{\K}{\X}$ be two polynomials. 
  If $\cm(p) \divleq \cm(q)$, 
  then $\lm(p) \divleq \lm(q)$, and $\dom(p) \subseteq \dom(q)$.
\end{lemma}
\begin{proof}
  obvious.
\end{proof}

Using this remark, one can show that the 
quasi-ordering $\pmonleq$ is well-founded whenever one assumes that
$(\mon[\omega]{\X}, \gdivleq)$ is a \kl{WQO}.

\begin{lemma}
  \label{lem:chm}
  The quasi-ordering $\pmonleq$ is \kl{compatible} with the action of $\group$,
  and is well-founded whenever one assumes that 
  $(\mon[\omega]{\X}, \gdivleq)$ is a \kl{WQO}.
\end{lemma}
\begin{proof}
  The first property is immediate because $\dom$, $\lm$, and $\revlexleq$ are
  compatible with the group action $\group$. 
  The second property follows from the fact that $\revlexlt$ is a total
  well-founded ordering whenever one has fixed finitely many possible 
  indeterminates. In a decreasing sequence, the support of the leading 
  monomials is also decreasing, so that sequence only contains finitely many 
  indeterminates, hence we conclude.
\end{proof}

As a consequence of \cref{lem:chm}, we know that the rewriting
relation $\toeucl{\pmonleq}{H}$ is \kl{terminating} for every set $H$. We can
also show that for a given polynomial, one can compute a finite set of
\emph{normalisations} of this polynomial with respect to $\Basis$, whenever 
$H$ is \kl{orbit-finite}.

\begin{definition}
  \label{def:normalisation}
  Let $H$ be a set of polynomials, and let $p \in \poly{\K}{\X}$ be a
  polynomial. We say that $p$ is \intro{normalised} with respect to $H$ if
  there are no transitions $p \toeucl{\pmonleq}{H} r$. 
  The set of \intro{remainders} of $p$ with respect to $H$ is 
  denoted $\rem{H}{p}$, and is defined as the set of all polynomials $r$ such that
  $p \toeucl{\pmonleq}{H}^* r$ and $r$ is normalised with respect to $H$.
\end{definition}

In order to compute in the equivariant setting, we will actually prefer to
manipulate \intro{orbit finite sets} of polynomials, that is, sets $H$ such
that there exists a finite set $H'$ of polynomials such that $H =
\orbit[\group]{H'}$, rather than talk about finite sets of polynomials and work
with their orbits directly.

\begin{lemma}
  \label{lem:normalisation}
  Let $H$ be an \kl{orbit finite set} of polynomials, and let $p \in \poly{\K}{\X}$ be a
  polynomial. Then $\rem{H}{p}$ is finite.
  Furthermore, this computation
  is equivariant is: for all $\gelem \in \group$, we have
  $\rem{H}{g \cdot p} = g \cdot \rem{H}{p}$.
\end{lemma}
\begin{proof}
  Let us write $H = \orbit[\group]{H'}$, where $H'$ is a finite set of
  polynomials.
  Because the relation $\toeucl{\pmonleq}{H}$ is \kl{terminating}, it suffices to 
  show that for every polynomial $p$, there are finitely many polynomials $r$ 
  such that $p \toeucl{\pmonleq}{H} r$. This is because 
  $p \toeucl{\pmonleq}{H} r$ implies that 
  $p = \alpha \monelt[n] (\gelem\cdot q) + r$ for some $q \in H'$, 
  $\alpha \in \K$, $\monelt[n] \in \mon{\X}$, and $\gelem \in \group$.
  Because, $\lm(r) \revlexlt \lm(q)$, we  
  conclude that $\lm(p) = \lm(\alpha \monelt[n] (\gelem\cdot q))$, and 
  therefore $r$ is uniquely determined by the choice of $q \in H'$ and the
  choice of $\gelem \in \group$ that maps the \kl{domain} of $q$ to the \kl{domain} of
  $p$. There are finitely elements in $H'$ and finitely many such functions
  $\gelem \in \group$ because both domains are finite.
\end{proof}

Now that we have a quasi-ordering on polynomials, we will prove that given a
finite set $H$ of generators, we can compute a \kl{weak equivariant Gröbner
basis}.
\begin{theorem}
  \label{thm:weakgb-comput}
  There exists an algorithm that takes as input a finite set $H$ of generators of an
  \kl{equivariant ideal} $\idl$ and computes a \kl{weak equivariant Gröbner
  basis} $\Basis$ of $\idl$ for the ordering $\pmonleq$.
\end{theorem}

\AP The computation will closely follow the classical \kl{Buchberger's
algorithm}. The main idea being to saturate the set of generators $H$ to remove
some \emph{critical pairs} of the rewriting relation $\toeucl{\pmonleq}{H}$.
Namely, given two polynomials $p$ and $q$ in $H$, we compute the set
$\CancelPoly{p}{q}$ of cancellations between $p$ and $q$ as the set of
polynomials of the form $r = \alpha \monelt[n] p + \beta \monelt[m] q$ such
that $\lm(r) < \max(\monelt[n] \lm(p), \monelt[m]\lm(q))$, where $\alpha,\beta \in \K$, and where
$\monelt[n], \monelt[m] \in \mon{\X}$. 

\begin{lemma}
  \label{lem:spoly}
  Let $p$ and $q$ be two polynomials in $\poly{\K}{\X}$.
  All the polynomials in $\CancelPoly{p}{q}$ are obtained by multiplying a monomial
  with
  the \kl{S-polynomial} $\spoly{p}{q}$ of $p$ and $q$, defined as
  \begin{equation}
    \label{eq:spoly}
    \spoly{p}{q} \defined
    \frac{\lcm(\lm(p), \lm(q))}{\lt(p)} \times p
    - \frac{\lcm(\lm(p), \lm(q))}{\lt(q)} \times q
    \quad .
  \end{equation}
\end{lemma}
\begin{proof}
  Let $p,q \in \poly{\K}{\X}$, and let $r \in \CancelPoly{p}{q}$.
  By definition, there exists $\alpha,\beta \in \K$ and $\monelt[n], \monelt[m]
  \in \mon{\X}$ such that $r = \alpha \monelt[n] p + \beta \monelt[m] q$ and
  $\lm(r) < \max(\monelt[n] \lm(p), \monelt[m] \lm(q))$.
  In particular,
  we conclude that $\lm(\monelt[n] p) = \lm(\monelt[m] q)$, and that 
  $\alpha \lc(\monelt[n] p) + \beta \lc(\monelt[m] q) = 0$.

  Let us write $\Delta = \lcm(\lm(p), \lm(q))$.
  Because $\lm(\monelt[n] p) = \lm(\monelt[m] q)$, there exists a monomial 
  $\monelt[l] \in \mon{\X}$ such that 
  $\lm(\monelt[n] p) = \monelt[l] \Delta = \lm(\monelt[m] q)$.
  Furthermore,
  we know that $\lc(p) \beta = - \lc(q) \alpha$.
  As a consequence, one can rewrite $r$ as follows:
  \begin{equation*}
    r = 
    \monelt[l] \alpha \lc(p) 
    \left[
      \frac{\Delta}{\lt(p)} \times p
      - \frac{\Delta}{\lt(q)} \times q
    \right]
    = 
    \monelt[l] \alpha \lc(p) \times \spoly{p}{q} \ .
  \end{equation*}
  We have concluded.
\end{proof}

Remark that the \kl{S-polynomial} is equivariant: if $g \in \group$, then
$\spoly{g \cdot p}{g \cdot q} = g \cdot \spoly{p}{q}$. We are now ready to
define our saturation algorithm. Whenever $p,q \in H$, these polynomials
clearly belong to $\IdlGen{H}$. Given a set $H$, we write $\spolyset(H)
\defined \bigcup_{p,q \in H} \rem{H}{\spoly{p}{q}}$.

\begin{lemma}
  \label{lem:spoly-orbit-finite}
  Let $H$ be an orbit finite set of polynomials. Then $\spolyset(H)$ is also an
  orbit finite set of polynomials.
\end{lemma}
\begin{proof}
  todo.
\end{proof}

We are now ready to define the saturation algorithm that will compute 
\kl{weak equivariant Gröbner bases}, described in \cref{alg:weakgb}.

\begin{algorithm}
    \caption{Computing \kl{weak equivariant Gröbner bases}}
    \label{alg:weakgb}
    \KwIn{An orbit finite set $H$ of polynomials}
    \KwOut{An orbit finite set $\Basis$ that is a \kl{weak equivariant Gröbner basis} of
      $\EqIdlGen{H}$}
    \Begin{
        $\Basis \gets H$\;
        \Repeat{$\Basis$ stabilizes}{
            $\Basis \gets \Basis \cup \spolyset(\Basis)$\;
        }
        \Return{$\Basis$}\;
    }
\end{algorithm}

In order for \cref{alg:weakgb} to be an actual algorithm, we need several 
effectivity assumptions. 

\begin{lemma}
  \label{lem:weakgb-computable}
  Assume that the order $\varleq$ is effectively computable, and 
  that the action of $\group$ is \kl{effectively oligomorphic}. Then, the
  algorithm \cref{alg:weakgb} is an actual algorithm,
  that can be run on a finite representation of the input set $H$.
\end{lemma}
\begin{proof}
  \todo[inline]{
    use the effectivity assumptions: one can compute the orbits of
    pairs of elements
    }
\end{proof}

\begin{lemma}
  \label{lem:weakgb-termination}
  Assume that $(\mon[\omega]{\X}, \gdivleq)$ is a \kl{WQO}. Then, 
  \cref{alg:weakgb} terminates.
\end{lemma}
\begin{proof}
  Let $\seqof{H_n}[n \in \N]$ be the sequence of sets of polynomials
  computed by \cref{alg:weakgb}. 
  We associate to each set $H_n$ the set $L_n$ of \kl{characteristic monomials} of the
  polynomials in $H_n$. Because the set of monomials is a \kl{WQO}, and because 
  the sequences are non-decreasing for inclusion, there exists an 
  $n \in \N$ such that, for every $\monelt \in L_{n+1}$, there exists
  $\monelt[n] \in L_n$, such that $\monelt[n] \gdivleq \monelt$.

  We will prove that $H_{n+1} = H_n$ by contradiction. Assume towards this
  contradiction that there exists some $r \in H_{n+1} \setminus H_n$. By
  definition of $H_{n+1}$, there exists $p,q \in H_n$ such that $r \in
  \rem{H_n}{\spoly{p}{q}}$. In particular, $r$ is \kl{normalised} with respect
  to $H_n$. However, because $r \in H_{n+1}$, $\cm(r) \in L_{n+1}$, and
  therefore there exists $\monelt[n] \in L_n$ such that $\monelt[n] \gdivleq
  \cm(r)$. This provides us with a polynomial $t \in H_n$ such that $\cm(t)$
  and an element $\gelem \in \group$ such that $\cm(t) \divleq \gelem \cdot
  \cm(r)$. Because $H_n$ is \kl{equivariant}, we can assume that $\gelem$ is
  the identity. In particular, one concludes that there exists $\monelt[n] \in
  \mon{\X}$ such that $\lm(t) \times \monelt[n] = \lm(r)$, and therefore
  $\lm(t) \gdivleq \cm(r)$. Therefore, one can find some $\alpha \in \K$ such
  that the polynomial $r' \defined r - \alpha \monelt[n] t$ satisfies $r'
  \pmonlt r$, and in particular, $r \toeucl{\pmonleq}{H_n} r'$.
  This contradicts the fact that $r$ is \kl{normalised} with respect to $H_n$.
\end{proof}

\begin{lemma}
  \label{lem:weakgb-correctness}
  Assume that $(\mon[\omega]{\X}, \gdivleq)$ is a \kl{WQO}. Then, the set $\Basis$ computed
  by \cref{alg:weakgb} 
  is a \kl{weak equivariant Gröbner basis} of the ideal
  $\EqIdlGen{H}$ for the ordering $\pmonleq$.
\end{lemma}
\begin{proof}
  It is clear that $\Basis$ is a generating set of $\EqIdlGen{H}$, because
  one only add polynomials that are in the ideal generated by $H$ at every step.

  Let $p \in \EqIdlGen{H}$ be a polynomial. Remark that 
  $p \in \IdlGen{\Basis}$ by the above argument and the fact that $\Basis$ is an
  \kl{equivariant} set.
  Let us write 
  \begin{equation}
    p = \sum_{i \in I} \alpha_i \monelt_i p_i
    \quad .
  \end{equation}
  Where $\alpha_i \in \K$, $p_i \in \Basis$, and $\monelt_i \in \mon{\X}$,
  for all $i \in I$.

  Let us assume without loss of generality that 
  the maximal leading monomial $\monelt[l] \defined \max(\seqof{\lm(\monelt_i p_i)}[i \in I])$ is
  minimal with respect to the ordering $\revlexleq$,
  which is a well-founded ordering thanks to \cref{lem:chm}.
  We distinguish two cases:

  \begin{description}
    \item[Case 1:] $\monelt[l] = \lm(p)$.
      In this case, one can remove the leading monomial $\monelt[l]$ from $p$ by
      substracting a suitable multiple of some element of its formal sum.
      In particular, the indeterminates of this are smaller than the ones
      of its leading monomial, hence smaller or equal to the indeterminates of $p$.

    \item[Case 2:] $\monelt[l] \neq \lm(p)$.
      In this case, it must be that the set $J$ the set of indices such that
      $\lm(\monelt_i p_i) = \monelt[l]$.
      Let us remark that 
      the sum of leading coefficients 
      of the polynomials in $J$ must vanish: $\sum_{i \in J} \alpha_i \lc(p_i) = 0$.
      As a concequence, the set $J$ has size at least $2$.
      Let us distinguish one element $\star \in J$, and 
      write $J_\star = J \setminus \set{\star}$.
      We conclude that 
      $\alpha_\star = - \sum_{i \in J_\star} \alpha_i \lc(p_i) / \lc(p_\star)$.
      Let us now rewrite $p$ as follows:
      \begin{equation}
        p = \sum_{i \in J_\star} \alpha_i 
        \left(\monelt_i p_i - \frac{\lc(p_i)}{\lc(p_\star)} \monelt_\star p_\star\right)
        + \sum_{i \in I \setminus J} \alpha_i \monelt_i p_i
        \quad .
      \end{equation}
      Now, by definition,
      polynomials $\alpha_i \monelt_i p_i$ for $i \in J \setminus J$ have 
      leading monomials
      strictly smaller than $\monelt[l]$.
      Furthermore,
      the polynomials
      $\monelt_i p_i - \frac{\lc(p_i)}{\lc(p_\star)} \monelt_\star p_\star$ for $i \in J_\star$
      cancel their leading monomials, hence they belong
      to the set $\CancelPoly{p_i}{p_\star}$.
      By \cref{lem:spoly}, we know that these polynomials are obtained by
      multiplying the \kl{S-polynomial} $\spoly{p_i}{p_\star}$ by some monomial.
      Because \cref{alg:weakgb} terminated, we know that 
      $\spoly{p_i}{p_\star} \toeucl{\pmonleq}{\Basis}^* 0$ by contruction.

      By definition of the rewriting relation, we conclude that 
      one can rewrite $\spoly{p_i}{p_\star}$ as combination of 
      polynomials in $\Basis$ that have smaller or equal leading 
      monomials. We conclude that
      the whole sum is composed of polynomials with leading monomials 
      strictly smaller than $\monelt[l]$, leading to a contradiction
      because we assumed $\monelt[l]$ to be minimal.
      \qedhere
  \end{description}
\end{proof}
